### 2023-05-03 paper.Rmd
editing

As a secondary purpose of Simulation 2, I investigated the degree to which partitions identified by the different methods showed similarity. Specifically, I investigated whether different anticlustering solutions showed any overlap regarding the placement of individual elements in groups. In cluster analysis, high quality partitions according to different objectives tend to overlap to a large degree, but it is an open question whether this result also holds for different anticlustering partitions. To tackle this issue, I computed the adjusted Rand index (ARI) for each pair of partitions that were returned by the competing methods [@rand1971; @hubert1985]. The ARI is an index of the similarity between two partitions, where an index of 0 indicates no similarity and a value of 1 indicates the highest possible similarity (i.e., when two partitions are the same). The pairwise ARIs were computed for each data set and for each $K$, i.e., 30,000 times.

@@

Table 3 illustrates the degree of overlap between anticlustering partitions that were identified in Simulation 2. Overall, it is shown that there was hardly any similarity between anticlustering partitions according to the different criteria. The average ARI was almost 0 in all cases---indicating no similarity of partitions whatsoever. This constitutes an interesting and novel result, illustrating that high between-group similarity can be achieved via very different partitionings.

@@

Anticlustering is usually implemented by reversing an objective function used in classical cluster analysis. Whereas cluster analysis seeks homogeneity within groups and separation between groups, the reversal results in heterogeneity within groups and similarity between groups. While tightly linked in a mathematical sense, an important semantic difference is that cluster analysis is concerned with finding "true" structure in the data, e.g., by defining a taxonomy on naturally occurring groups [@everitt1979]. Anticlustering instead imposes structure on the data, while the anticlusters themselves do not represent any underlying "true" pattern. Correspondingly, different---but similarly adequate---anticlustering partitions tend to look very different, as shown in Simulation 2. In contrast, good clustering solutions tend to look similar because a good clustering is characterized by assigning similar elements into the same cluster and dissimilar elements into different clusters. A good anticlustering solution however does not depend on assigning individual items either to the same or a different group, independent of their pairwise dissimilarity.[^dispersion] This distinction can also be clarified using Figure 1, where we observe a high concentration of suitable anticlustering partitions but only few suitable clustering solutions. An astonishing `r mean(df$kmeans / max(df$kmeans) >.999) * 100`% of all partitions are within 0.1% of the global maximum in $\mathit{SSE}$, i.e., the best k-means anticlustering solution. In contrast, there are only `r sum(min(df$kmeans)/ df$kmeans > .90)` partitions (`r mean(min(df$kmeans)/ df$kmeans > .90)*100`%) that approximate the global minimum in $\mathit{SSE}$ (i.e., the best k-means clustering solution) by even 10%. Thus, even though the problems are based on the same objective function, cluster analysis and anticlustering exhibit a stark discrepancy with regard to the solution space of partitions. Interestingly, the k-plus objective introduces an additional distinction between cluster analysis and anticlustering. Reversing the k-plus anticlustering objective no longer yields a useful clustering interpretation, because identifying groups that are different with regard to higher order moments does not constitute a clustering. Thus, even though k-plus extends the classical k-means clustering criterion, it is a pure anticlustering objective.

>>>>

As a secondary purpose of Simulation 2, I investigated the degree to which partitions identified by the different anticlustering methods showed similarity. Specifically, I investigated whether different anticlustering solutions showed any overlap regarding the placement of individual elements in groups. In cluster analysis, high quality partitions according to different objectives tend to overlap to a large degree, but it is an open question whether this result also holds for different anticlustering partitions. To tackle this issue, I computed the adjusted Rand index (ARI) for each pair of partitions that were returned by the competing methods [@rand1971; @hubert1985]. The ARI is an index of the similarity between two partitions, where an index of 0 indicates no similarity and a value of 1 indicates the highest possible similarity (i.e., when two partitions are the same). The pairwise ARIs were computed for each data set and for each $K$, i.e., 30,000 times.

@@

Table 3 illustrates the degree of overlap between anticlustering partitions that were identified in Simulation 2. Overall, it is shown that there was hardly any similarity between anticlustering partitions according to the different criteria. The average ARI was almost 0 in all cases---indicating no similarity of partitions whatsoever. Moreover, the ARI between two anticlustering partitions hardly differed from the ARI between an anticlustering partition and a random partition. This constitutes an interesting and novel result, illustrating that high between-group similarity can be achieved via very different partitionings.

@@

Anticlustering is usually implemented by reversing an objective function used in classical cluster analysis. Whereas cluster analysis seeks homogeneity within groups and separation between groups, the reversal results in heterogeneity within groups and similarity between groups. While tightly linked in a mathematical sense, an important semantic difference is that cluster analysis is concerned with finding "true" structure in the data, e.g., by defining a taxonomy on naturally occurring groups [@everitt1979]. Anticlustering instead imposes structure on the data, while the anticlusters themselves do not represent any underlying "true" pattern. Correspondingly, different---but similarly adequate---anticlustering partitions tend to look very different, as was also shown in Simulation 2. In contrast, different clustering solutions tend to look similar because a good clustering is characterized by assigning similar elements into the same cluster and dissimilar elements into different clusters. A good anticlustering solution however does not depend on assigning individual items either to the same or a different group, independent of their pairwise dissimilarity.[^dispersion] This distinction can also be clarified using Figure 1, where we observe a high concentration of suitable anticlustering partitions but only few suitable clustering solutions. An astonishing `r mean(df$kmeans / max(df$kmeans) >.999) * 100`% of all partitions are within 0.1% of the global maximum in $\mathit{SSE}$, i.e., the best k-means anticlustering solution. In contrast, there are only `r sum(min(df$kmeans)/ df$kmeans > .90)` partitions (`r mean(min(df$kmeans)/ df$kmeans > .90)*100`%) that approximate the global minimum in $\mathit{SSE}$ (i.e., the best k-means clustering solution) by even 10%. Thus, even though the problems are based on the same objective function, cluster analysis and anticlustering exhibit a stark discrepancy with regard to the solution space of partitions. Interestingly, the k-plus objective introduces an additional distinction between cluster analysis and anticlustering. Reversing the k-plus anticlustering objective no longer yields a useful clustering interpretation, because identifying groups that are different with regard to higher order moments does not constitute a clustering. Thus, even though k-plus extends the classical k-means clustering criterion, it is a pure anticlustering objective.
